---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
library(ggplot2)
library(plyr)

activities <- read.csv("activity.csv")
activities$day <- weekdays(as.Date(activities$date))
activities$DateTime<- as.POSIXct(activities$date, format="%Y-%m-%d")
```

### Removing NAs from the data 
```{r}
activities_clean <- activities[!is.na(activities$steps),]
```
  

## What is mean total number of steps taken per day?

### 1. summarizing total steps per date
```{r}
sumTable <- aggregate(activities_clean$steps ~ activities_clean$date, FUN=sum)
colnames(sumTable)<- c("Date", "Steps")
print(sumTable)
```
### 2. Creating the historgram of total steps per day
```{r histogram, fig.height=4}
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day")
```

```{r echo=FALSE,results='hide'}
meansteps <- as.integer(mean(sumTable$Steps))
mediansteps <- as.integer(median(sumTable$Steps))
```
### 3. Mean and Median of steps
The average number of steps taken each day was `r meansteps` steps.

The median number of steps taken each day was `r mediansteps` steps.


## What is the average daily activity pattern?

### create average number of steps per interval
```{r}
act_intervals <- ddply(activities_clean, .(interval), summarize, Avg=mean(steps))
```
### Create line plot of average number of steps per interval
```{r}
p <- ggplot(act_intervals, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
##Maximum steps by interval
maxSteps <- max(act_intervals$Avg)
##Which interval contains the maximum average number of steps
maxinterval <- act_intervals[act_intervals$Avg==maxSteps,1]
maxSteps <- round(maxSteps)
```
The maximum number of steps for a 5-minute interval was `r maxSteps` steps.

The 5-minute interval which had the maximum number of steps was the `r maxinterval` interval.


## Imputing missing values

### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
##Number of NAs in original data set
na_rows <- nrow(activities[is.na(activities$steps),])
```
The total number of rows with steps = 'NA' is `r na_rows`.

### 2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r}
## Create the average number of steps per weekday and interval
avgTable <- ddply(activities_clean, .(interval, day), summarize, Avg=mean(steps))

## Create dataset with all NAs for substitution
nadata<- activities[is.na(activities$steps),]

## Merge NA data with average weekday interval for substitution
newdata<-merge(nadata, avgTable, by=c("interval", "day"))
```

### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
## Reorder the new substituded data in the same format as clean data set
newdata2<- newdata[,c(6,4,1,2,5)]
colnames(newdata2)<- c("steps", "date", "interval", "day", "DateTime")

##Merge the NA averages and non NA data together
mergeData <- rbind(activities_clean, newdata2)
```

### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
##Create sum of steps per date to compare with step 1
sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum)
colnames(sumTable2)<- c("Date", "Steps")

## Mean of Steps with NA data taken care of
newmean <- as.integer(mean(sumTable2$Steps))

## Median of Steps with NA data taken care of
newmedian <- as.integer(median(sumTable2$Steps))

## Difference between new mean/median and old mean/median before imputing
diffmean <- newmean - meansteps
diffmedian <- newmedian - mediansteps


## Creating the histogram of total steps per day, categorized by data set to show impact
hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Black")
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Grey", add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("black", "grey") )


```

The new mean of the imputed data is `r newmean` steps compared to the old mean of `r meansteps` steps. That creates a difference of `r diffmean` steps on average per day.

The new median of the imputed data is `r newmedian` steps compared to the old median of `r mediansteps` steps. That creates a difference of `r diffmedian` steps for the median.

## Are there differences in activity patterns between weekdays and weekends?

### 1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
## Create new category based on the days of the week
mergeData$DayCategory <- ifelse(mergeData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
```

### 2. Make a panel plot containing a time series plot (i.e. ) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
library(lattice) 

## Summarize data by interval and type of day
intervalTable2 <- ddply(mergeData, .(interval, DayCategory), summarize, Avg = mean(steps))

##Plot data in a panel plot
xyplot(Avg~interval|DayCategory, data=intervalTable2, type="l",  layout = c(1,2),
       main="Average Steps per Interval Based on Type of Day", 
       ylab="Average Number of Steps", xlab="Interval")
```


